Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-13 Manuscript/MPESubmission/Revision/response.tex

\begin{enumerate}[(1)]
\item The Total Evidence method places living and fossil taxa on the same phylogeny.
\item Here we test the effect of missing morphological data on Total Evidence topologies.
\item The number of living taxa with morphological data has greatest effects on topology.
\item Bayesian inference outperforms Maximum Likelihood for correct clade recovery.
\item We recommend increased sampling effort for morphological data of living taxa.
\end{enumerate}

% -----------------
% Comment 4
% -----------------
\item{\textcolor{blue}{\textbf{Introduction:} Needs stronger emphasis on what specifically is unique about this study compared to what the many other studies on missing data have done and found.
Clearly outlining the questions in this study that are left open from previous research is important because some of the findings in this study suggest that there may be more error associated with missing data than previous research has suggested.}}

We have increased our emphasis on novelty throughout the introduction by more clearly stating the difference between the present study and the literature and more clearly outlining the question of this study, as follows:

\textit{Until now, few attempts have been made to study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework (i.e. using both molecular and morphological data; e.g. Wiens et al., 2005; Manos et al., 2007; Pattinson et al., 2014).
These previous studies assessed the effect of missing data on topology by either (1) comparing a dataset with missing data to subsets without missing data (Wiens et al. 2005); or (2) removing both molecular and some morphological data from living taxa to create artificial fossils (Manos et al., 2007; Pattinson et al., 2014).
Both approaches have shown that missing data are not a major problem and should not be an obstacle to combining both living and fossil species in the same phylogenies.
The way these studies were conducted, however, means that their conclusions are not generally applicable across all scenarios involving missing data in Total Evidence phylogenies.
For example, using an empirical (rather than simulation based) approach limits their conclusions to studies with similar distributions of data across species in the phylogeny.
Additionally, Wiens et al. (2005) did not include fossil taxa in their analyses, so their results cannot be used to make conclusions about how missing data may influence the placement of fossils. 
Manos et al. (2007) and Pattinson et al. (2014) include fossil taxa, but used the patchiness of the fossil record to determine how to remove data from their matrices. 
Data for living species are unlikely to be missing in this patchy way, instead full molecular data with the complete absence of morphological data is a likely pattern (Guillerme and Cooper 2015).
Finally, Manos et al. (2007) and Pattinson et al. (2014), mainly focused on how missing data in fossil taxa affect the placement of fossils, ignoring the effects of missing data in living species.}

\textit{In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by removing living taxa with morphological data, fossil data, all data for certain characters and the combination of these three aspects.
This is an advance on previous studies because we use large-scale simulations and analyze the effects of three distinct aspects of missing data thus focusing on both neontological and paleontological parts of the matrix.
In addition, we test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian phylogenies: (i) the conservation of clades (based on the Robinson-Foulds distance; Robinson and Foulds, 1981) and (ii) the displacement of wild-card taxa (based on the Triplets distance; Critchlow et al., 1996) rather than just a single measure of clade conservation or clade support (cf. Wiens et al., 2005; Pattinson et al., 2014).} lines 85-118

@@

Although though we agree with the comments of both reviewers that knowing the difference between the true and ``best''/``missing-data'' trees is important, we do not think that these results warrant inclusion in the main text.
In these simulations, we had to make a choice between (1) simulating larger matrices but with fewer chains (i.e. less variance) or (2) smaller matrices with more variance.
We deliberately made the second choice because our main question was "what is the effect of our parameters on recovering topology".
To answer this question, we maintain that comparing the best tree to the missing-data trees is the best approach for two reasons:
\begin{enumerate}
\item{The true tree is always unknown:} as we state in the manuscript: "In real world situations, the ``true'' tree is never available" (line 242 in the first manuscript and @@@ in the revision).
\item{The true tree is just the starting point of the simulations:} we use the true tree only as a starting point for our simulations.
Once generated, the true tree is fed to a cascade of algorithms and software (i.e. seqGen\{phyclust\}, rTraitDisc\{ape\}, RAxML, MrBayes) described in section 2.1 and 2.3 to get the best tree and the missing-data trees.
Comparing the latter to the true tree will answer our question "what is the effect of our missing-data parameters and the cascade of algorithms and software on recovering the true topology?".
\end{enumerate}
Therefore we added a second justification to the main text (see below this response) and added these analysis (``true'' \textit{vs.} ``best''/``missing-data'' trees) to the supplementary for readers that might be interested by this specific problem.

We added the following sentence to the main text to clarify and link to the supplementary analysis:

\textit{We compare ``best'' trees to ``missing data'' trees but could also compare ``true'' trees to the ``missing data'' trees.
In practice, the difference between the ``best'' trees and the ``missing data'' trees represents the effect of our missing data parameters and of the phylogenetic methods used to infer the ``missing data'' trees.
The difference between the ``true'' and the ``missing data'' trees, however, represents the effect of our parameters used to generate the ``true'' tree and the algorithms used to generate the ``complete'' matrix as well as the effect of our missing data parameters and the phylogenetic methods used.
Because the main aim of this study is to look at the effect our missing data parameters on topological recovery, we chose to represent only the comparisons between the ``best'' trees and ``missing data'' trees.
The results of the comparisons of the ``true'' tree and the ``missing data'' trees are available in Appendix B.
Note that this makes little difference to our overall results.} line 297-308 @@@.

@@

The difference between the Bayesian consensus and the Bayesian posterior trees can be due to two factors:
\begin{enumerate}[I]
\item the way we assess topological recovery is slightly different in each case: for the Bayesian consensus trees, we simply compared the missing-data consensus tree to the best consensus tree; for the Bayesian posterior trees however, we randomly compared 1000 missing-data Bayesian posterior trees to 1000 best Bayesian posterior trees.
Therefore, even if the comparison methodology is the same, the fact that the second methods draws random trees means we expect the topological recovery score to be blurred in comparison to other comparisons that do not involve random comparisons.
\item the Bayesian posterior trees are always resolved and thus, in the same way as the Maximum Likelihood tree, more likely to contain false-positive nodes. On the other hand, because the consensus tree is a strict majority consensus rule, only nodes with a a good support will be represented in the tree and all other nodes will be collapsed. This reduces the number of false-positive nodes (since an unresolved tree is always better than a wrongly resolved one) and increases clade conservation but also increases the displacement of wild-card taxa (since it is likely that some Triplets will just not exist in a poorly-resolved consensus tree).
\end{enumerate}
Both points were discussed in the original manuscript lines 566-573 (or lines 645-652 @@@ in the revised version).
We expanded the paragraph to:

\textit{Additionally, the Bayesian posterior trees performed more poorly than the Bayesian consensus tree (Fig. 4, Table 1 and Appendix C Fig. C5 and Tables C5, C6 and C7).
This may be because the Bayesian posterior trees are always resolved and thus more likely to contain incorrectly resolved nodes (i.e. decreasing the Normalised Robinson-Foulds metric).
Conversely, the Bayesian consensus trees might not resolve nodes that are poorly supported and thus are more likely to contain only correctly resolved nodes (i.e. increasing the Normalised Robinson-Foulds metric).} lines 681-687. @@@

@@

% -----------------
% Comment 10
% -----------------
\item{\textcolor{blue}{\textbf{Results - Bayesian Triplets:} Another important result that is not mentioned is that using the wildcard taxon metric (Tr), Bayesian consensus trees do quite poorly with missing data.
Values in Figures 4 and 5 clearly drop below zero, although negative values are not illustrated on the axis, and this indicates that the tree topologies are no better than, or even worse than, random trees.
It is especially true when fossils are missing data, and that is going to be an important finding with practical application but must be addressed. }}

We modified the Figs. 4 and 5 to show Normalised Triplets metric values below 0 and added a dashed line at 0 to underline the difference in scale between both metrics.
As mentioned above (reviewer 1 comment 9), this is due to the fact that the Bayesian consensus tree will be less and less resolved when missing data increases.
This has two related effects.
The first one being to retain "good" clade conservation (i.e. avoiding false-positive nodes by collapsing unsupported ones).
And the second one is that when these clades are collapsed, it reduces the number of available triplets and thus increases the wild-card taxa score (Normalised Triplets metric).
In fact, when a clade ((a;b);c); is collapsed to (a;b;c), the whole clade is conserved (a, b and c are still in a common clade) however they also all become wild-card taxa: the position of a regarding b and c is now incorrect.

We added a note on this in the section about the effects of the tree inference method:

\textit{The Bayesian consensus trees, however, perform poorly for the Normalised Triplets metric: some parameter combinations, especially when the $M_F$ parameter reaches 75\% missing data, lead to negative values (Fig. 5).
A Normalised Triplets metric value below 0 means that the placement of some taxa is worse than expected by just randomly placing this taxon in the tree.
This can be interpreted as the absence of comparable triplets between some of the ``missing data" trees and ``best'' trees.
Even if clades are conserved (Fig. 5), the resolution within them can be poor to non-existent when a large proportion of data are missing (i.e. 75\%).
In such cases, the fossil taxa are equally likely to be placed in any of the clades that they share the most characters with.
These results are in agreement with previous studies that have showed that missing data can cause problems for recovering ``correct'' topologies, especially for small matrices of 100 characters (Wiens 2003).
It is important to note, however, that this effect can be reduced by increasing the number of characters (Wiens 2003).} lines 661-673. @@@

@@

\textit{Previous studies have explored the effect of missing morphological data in Total Evidence matrices (Wiens et al., 2005; Manos et al.,2007, Pattinson et al., 2014).
The conclusions of theses studies, however, were limited by their empirical approach making their results applicable only to similar missing data scenarios.
Additionally, these studies focused either only on living taxa (Wiens et al., 2005) or on the patterns of missing data from the fossil record only (Manos et al., 2007, Pattinson et al., 2014).
Here we instead used an approach where missing data were generated from simulated data and according to three clearly defined missing-data parameters ($M_{L}$, $M_{F}$ or $N_{C}$) that removed data from both the living and fossil taxa.
This allowed us to confirm previous results that missing data can be especially problematic in small matrices (Wiens 2003), but also revealed the crucial importance of coding morphological data for living species in Total Evidence phylogenies.
Missing data in Total Evidence matrices is not a problem for recovering the ``best'' tree topology as long as enough living and fossil taxa in the matrix have data for overlapping morphological characters.
When missing data increases in any of our missing data parameters ($M_{L}$, $M_{F}$ or $N_{C}$), it reduces support for the placement of fossil taxa and increases the displacement of wildcard taxa.
Therefore we advise increased focus on coding morphological characters for a large number of the living taxa present in the matrix (i.e. at least 50\%) if the goal is to accurately combine both living and fossil species in phylogenies.
Doing so will increase overlap of morphological characters among living and fossil taxa, allowing the fossil taxa to be positioned relative to the living taxa based on their shared derived characters rather than simply on available data.}

\textit{Additionally, the topologies of the Bayesian consensus trees, regardless of the amount of missing data, were always closer to the ``best'' tree topology than the Maximum Likelihood trees.
This has also been observed in empirical data (e.g. Arcila et al., 2015) where Maximum Likelihood trees inferred from a Total Evidence matrix were less supported than the Bayesian consensus tree.
This might have an important impact on estimating topologies in the Total Evidence framework, because previous studies had to rely either on molecular scaffolds (e.g. Slater, 2013), taxonomic constraints (e.g. Slater, 2013; Beck and Lee, 2014) or even on fixing the topology (e.g. Ronquist et al., 2012a).
Therefore, we suggest extracting such topological backbones from the Bayesian consensus tree if needed.} lines 731-62@@@

@@

We changed the axes of the graphs showing Normalised Triplets metric to show values below 0 throughout the manuscript and the appendices.

% -----------------
% Comment 14
% -----------------
\item{\textcolor{blue}{\textbf{Table caption:} Table caption should include a line about 'pooling across missing data schemes' to clarify what is presented.}}

We clarified the caption of the Table 1 as follows:

\textit{Each line summarizes the probabilities of overlap between the distributions of the ``best'' tree versus trees from each inference method (Maximum Likelihood; Bayesian consensus; Maximum Likelihood Bootstraps and Bayesian posterior trees) pooled across all combinations of missing data parameter values, using the Normalised Robinson-Foulds (RF) and Triplets (Tr) metrics. 
Values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the ``correct'' tree topology. Values $>$ 0.95 denote significantly similar distributions and values $<$ 0.05 denote significantly different distributions.}

@@

\end{enumerate}

\section{Reviewer 1 Specific comments:}
\begin{enumerate}

% -----------------
% Specific Comment 1
% -----------------
\item{\textcolor{blue}{\textbf{line 31:} has it really been shown that the majority of macroevolutionary studies focuses on extant species only? The cited references are examples of large scale extant-only phylogenies. To make such a pointed statement, the authors would need to find support from some kind of literature review or meta-analysis comparing the number of paleontological to neontological macroevolutionary studies. Otherwise, rephrase. }}

We revised this statement as follows: \textit{many large-scale macroevolutionary studies focus solely on living species} (line 30-31).

@@

% -----------------
% Specific Comment 46
% -----------------
\item{\textcolor{blue}{\textbf{line 549-565:} again, I am left wondering how much the differences in performance of max. likelihood and Bayesian inference techniques are driven by the use of the 'true' tree as a starting tree in the Bayesian MCMC search, while the likelihood analysis had no such guide tree to begin with. The authors at least need to make this point clear, if not re-run some subset of analyses without introducing a starting tree to test the effects of using the 'true' tree as a starting tree or not. I understand the computational limitations, but the practical limitation that empiricists will never have the 'true' tree to start with may (or may not) have a large impact on our interpretation and the generality of these findings. }}

As discussed above (reviewer 1, specific comment 29), we understand the reviewer's concern and have added an analysis in Appendix A (Fig. A2 and Tables A1 and A2) where we tested the effect of using the ``true'' tree as a starting tree. We found that there is no significant effect of using the ``true'' tree instead of a random tree (MrBayes default) as a starting tree on topological recovery.

We refer to this supplementary analysis in the main text as follows:

\textit{Our results show that the topology of the Bayesian consensus tree is always closer to the ``best'' tree topology than the ``best'' Maximum Likelihood tree (Fig. 5).
Note that the methodological choice of using the ``true'' tree as a starting tree for the Bayesian Inference rather than a random starting tree (see Methods), had no significant effect on topological recovery (see Appendix A, section ``Effect of the starting tree on Bayesian inference'' for details).} lines 649-655@@@

@@

% -----------------
% Specific Comment 48
% -----------------
\item{\textcolor{blue}{\textbf{line 599:} remove extra close parenthesis inside parenthetical statement}}

We fixed this typo.

% -----------------
% Specific Comment 49
% -----------------
\item{\textcolor{blue}{\textbf{line 595-606:} this paragraph seems unsatisfactory advice for practitioners.
First, advising that the tree topology be fixed using the Bayesian consensus tree before conducting analyses such as dating strongly conditions the results on only a single tree and as the authors point out in the parenthetical statement, the dating information could improve accuracy. Indeed one of the many advantages of tip-dating is the joint inference of tree topology and divergence times.
The following statements that the posterior distribution should not be discarded does not leave the reader with a clear path that they should follow - if only the fixed tree topology is used to estimate divergence times, how can the substitution rate-scaled posterior distribution of highly probable trees be used for most comparative analyses which require time-scaled branch lengths?
This paragraph requires further justification and this study does have important bearing on a subject that gets little attention. }}

We revised the paragraph as follows:
\textit{Another practical implication of our results regards the tree inference methods.
Because the Bayesian consensus trees consistently recovered topologies closer to the ``best'' tree topology than the Maximum Likelihood trees, we advise that where a topological constraint is needed, Bayesian consensus trees should be used. This may apply to tree inferences using the Total Evidence method such as tip-dating (e.g. Ronquist et al., 2012a; Wood et al., 2013; Matzke, 2014). It is, however, possible that including dating information during tree inference could also improve the accuracy of the Bayesian posterior tree distribution, so a fixed topology should be used with caution.
Using the Bayesian consensus tree rather than the Maximum Likelihood can also reduce the number of false positive topologies (\textit{sensu} Swofford et al., 2001).
As shown in Fig. 5 and discussed in the section above (Effects of tree inference methods), the Bayesian consensus tree is more likely to not resolve poorly supported nodes due to missing data than the Maximum Likelihood tree that is more likely to incorrectly resolve such nodes (i.e. creating a false positive node).
Note, however, that we do not suggest discarding the Bayesian posterior tree distributions even though they performed poorly in recovering the ``best'' tree topology in our simulations (this can probably be traced to the difficulties comparing distributions of trees; see above).
These trees will be invaluable for phylogenetic comparative analyses. For example a sub-sample of posterior tree distributions can be used to assess macroecological questions while better taking into account topological uncertainty (e.g. Fritz et al. 2009 and Jetz et al. 2012 trees used in Healy et al. 2014).} lines 711-731@@@

@@

% -----------------
% Comment 2
% -----------------
\item{\textcolor{blue}{\textbf{Better explanation of the missing data:} I found the explanation of the missing data strategy very difficult to follow. First, they oddly use the verb "to infer" for matrices; are they inferring matrices? Second, it's not clear anywhere that the authors actually removed the molecular data from the fossil taxa, though I assume they did. I hope the authors can re-write in simpler language ("explain like I'm 5") for easier comprehension.}}

The lines 180 and 181 states "All the molecular information for fossil taxa was replaced by missing data ("?")" in the former manuscript. We changed the verb "infer" (see specific comment number 16 from reviewer 1).

We assume the reviewer is referring to section "2.2 Removing data". We therefore modified the whole section to improve its clarity and simplicity as follows:

### 2015-08-13 Manuscript/MPESubmission/Revision/response.tex

\textit{To explore the effect of missing morphological data on topological recovery, we removed various amounts of the ``complete'' matrix to obtain matrices with missing morphological data.
Hereafter, we call these matrices with missing morphological data the ``missing-data'' matrices.
Note that the amount of molecular data remained constant throughout our simulations: 1000 molecular characters for living taxa and no molecular data for fossil taxa (see above).
We removed morphological data using three data incompleteness parameters:
\begin{enumerate}
\item{The proportion of missing living taxa ($M_L$).} This first missing-data parameter corresponds to the proportion of living taxa with no morphological data.
It represents the number of living taxa that are present in the matrix but have only molecular data available.
This reflects the fact that, because of the increasing ease of collecting molecular data, morphological data for living species are rarely collected (Guillerme and Cooper, 2015).
Therefore, many living species will have only molecular data available.
In practice, we removed all the morphological data from randomly chosen living taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of living taxa with no morphological data.
\item{The proportion of missing data in the fossil record ($M_F$).} This missing data parameter represents the completeness of the fossil record. Due to preservation biases, missing data for fossil taxa are common (Sansom and Wills, 2013).
In practice, we randomly removed a proportion of data from across the fossil taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of overall missing data for the fossil taxa.
Note that 50\% missing data for fossil taxa does not mean that each fossil is missing 50\% of its morphological data.
Instead this 50\% refers to missing fossil data across the whole matrix.
Some fossils may retain 100\% of their data and others may lose most of their data at this parameter value (down to a minimum threshold of 5\% available data; see below)
\item{The number of morphological characters for both living and fossil taxa ($N_C$).}
This parameter is not a missing data parameter \textit{per se} but rather an indication of the size of the matrix.
Any morphological matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large.
Therefore, this parameter corresponds to the overall number of characters available for both living and fossil taxa.
In practice, we randomly removed entire characters from the morphological matrix reducing it to: 100, 90,75, 50 or 25 characters.
Note that these levels are equivalent to the two other parameters (i.e. 0\%, 10\%, 25\%, 50\% or 75\% of ``missing'' morphological characters).
\end{enumerate}}

### 2015-08-13 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
Matched response and manuscript

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices.
Since the molecular part of a Total Evidence matrix acts like a ``classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}.
Therefore, we focus only on missing data in the morphological part of the matrix.
We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the proportion of missing morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix).
%in
\hl{We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the resulting tree topology.
We infer the topology from the matrices using both Maximum Likelihood and Bayesian inference methods and measure the differences in topology using two different topological distance metrics as proxies for clade conservation } \citep[\hl{based on the Robinson-Foulds distance;}][]{RF1981}\hl{ and for wild-card taxa placement }\citep[\hl{based on the Triplets distance;}][]{critchlowthe1996}.
%\hl{Thus, in the present study, we propose a new approach to the missing-data question, focusing on both palaeontological and neontological data (rather than just palaeontological data), using only simulated data and measuring two aspects of topological recovery }\citep[cf.][\hl{measuring respectively only the clades support or the number of shared clades}]{Wiens01102005,pattinsonphylogeny2014}.

% NC: Reading this in full it seems like the last bit repeats what you've already said. I know the reviewer asked for novelty to be spelled out but I think you do that well above here so maybe leave this bit out to avoid repetition.

% NC: You've also cut out the "here we find" sentences. Did the reviewers ask for this? If not why not put them back and mention novelty in there instead subtly.

%Former revised version
% \hl{We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods by using two different topological metrics, based on the Robinson-Foulds }\citep{RF1981}\hl{ and Triplets }\citep{critchlowthe1996}\hl{ distances.
% Thus, in the present study, we propose a new approach on the missing-data question focus on both palaeontological and neontological data (rather than just palaeontological data), using only simulated data and two distinct topological metrics.}
% We find that minimizing the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the ``best'' tree topology more so than minimizing the amount of missing data in the fossil record.
% Additionally, we find that the ability of Total Evidence methods to recover the ``best'' tree topology is increased when using Bayesian methods.

>>>>

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices.
Since the molecular part of a Total Evidence matrix acts like a ``classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}.
Therefore, we focus only on missing data in the morphological part of the matrix.
We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the amount of morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix).
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the resulting tree topology.
We infer the topology from the matrices using both Maximum Likelihood and Bayesian inference methods and measure the differences in topology using two different topological distance metrics as proxies for clade conservation and for wild-card taxa placement.
We find that minimizing the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the ``best'' tree topology more so than minimizing the amount of missing data in the fossil record.
Additionally, we find that the ability of Total Evidence methods to recover the ``best'' tree topology is increased when using Bayesian methods.
